---
title: "Further exploratory data analysis"
author: "Charles"
date: "Last compiled on `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    theme: cosmo
    highlight: monochrome
    toc: yes
    toc_float: no
    toc_depth: 6
    code_folding: hide
  word_document:
    toc: yes
    toc_depth: '6'
  pdf_document:
    toc: yes
    toc_depth: '6'
editor_options:
  chunk_output_type: console
---

```{css, echo=FALSE}
<style>
body, h1, h2, h3, h4 {
    font-family: "Bookman", serif;
}

body {
    color: #333333;
}
a, a:hover {
    color: red;
}
pre {
    font-size: 10px;
}
</style>
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Introduction

In this script I perform further exploratory data analyses on DNA methylation data from TGCA

## Load libraries

```{r results='hide'}
library(coloncancermeth)
data(coloncancermeth)
dim(meth)
dim(pd)
```


## Exploratory data analysis

### Use the methylation profiles to compute a Euclidean distance between each sample.

```{r results='hide'}
d <- dist(t(meth))
```

Now use **cmdscale** function.

```{r fig.align='center'}
mds <- cmdscale(d)
colnames(mds) <- c("PC1", "PC2")

mypar(1,1)
plot(mds[, "PC1"],
     mds[, "PC2"],
     pch=21,bg=as.numeric(pd$Status),
     xlab = "First dimension",
     ylab = "Second dimension")
legend("bottomright",levels(pd$Status),
       col=seq(along=levels(pd$Status)),pch=15,cex=0.8)
```

The MDS plot shows separation between cancer and normal samples, but only in the first dimension. The second dimension seems to be associated with a large variability within the cancers. 

### For each CpG compute p-values for the cancer versus normal comparison using the limma package.

```{r}
library(limma)
X <- model.matrix(~pd$Status)
fit <- lmFit(meth, X)
eb <- eBayes(fit)
pvals <- eb$p.value[, 2]
q_values <- qvalue::qvalue(pvals)
mean(q_values$qvalues < 0.05)
```

Only 23.73% of CpGs are differentially methylated in tumour vs healthy conditions based on their q values (corrected for multiple hypothesis testing).

```{r}
Index <- which(q_values$qvalues <= 0.05)
sig_coefs <- eb$coefficients[Index, ]
mean(sig_coefs[, 2] > 0)
```

41% of CpGs that are significantly differentially methylated are more methylated in the tumour compared to healthy samples.

### Determine which of the differentially methylated CpGs are in CpG islands.

```{r}
library(AnnotationHub)
ah <- AnnotationHub()
cgi <- ah[["AH5086"]] # islands
islands=gr[Index]%over%cgi
hypermethylated=fit$coef[Index,2]>0
prop.table( table(islands,hypermethylated) )
```

26.26% of differentially methylated CpGs are in CpG islands.

### Now we will use the bumphunter package to separate the differentially methylated CpGs intro groups.

```{r}
library(bumphunter)
X <- model.matrix(~pd$Status)
chr <- as.character(seqnames(gr))
res <- bumphunter(meth,
                  X,
                  chr = chr,
                  pos = start(gr),
                  cutoff = 0.1)
head(res$table)

# filtering by zone
dmrs <- res$table[ res$table$L>=3, ]
dmrs <- makeGRangesFromDataFrame(dmrs)
dist_to_island <- distanceToNearest(dmrs, cgi)
mean(dmrs %over%cgi)
```

61.05% of DMRs overlap a CpG island and `r mean(d>0 & d<=2000) * 100` % are within 2000 bp from a CpG island but do not overlap.

```{r}
sessioninfo()
```

